Incorporating periodic variability in hidden Markov models for animal movement

### 2016-02-29 paper3.Rnw
latex compiling problem

\doublespacing
<<pkgs,message=FALSE,warning=FALSE,echo=FALSE>>=
library("ggplot2"); theme_set(theme_bw())
scale_colour_discrete <- function(...,palette="Set1")
    scale_colour_brewer(...,palette=palette)
scale_fill_discrete <- function(...,palette="Set1")
    scale_fill_brewer(...,palette=palette)
zmargin <- theme(panel.margin=grid::unit(0,"lines"))
library("knitr")
opts_chunk$set(fig.width=6,fig.height=4)
library("depmixS4")
library("plyr")
library("dplyr")
library("reshape2")
@

@@

Figure 1 shows the BIC-optimal number of states for time homogenous models is consistent with van de Kerk et al. Furthermore, BIC-optimal number of states does not depend on the model for both time-homogenous and time-inhomogenous within their respective framework, but time-inhomogenous models have the BIC-optimal number of states overall.

@@

As a complement, we also added FMM, FMM with sinusoidal temporal heterogeneities, and HMM with hourly transitions to compare the temporal effects in goodness of fit. Looking at Figure 2, time inhomogenous transition HMMs (vertically) and 5 states yield the lowest $\Delta$ BIC. The time inhomogeneous transition HMMs reduced the number of states from the HMM by 1 state consistently across all panthers via $\Delta$ BIC. Two of the four panthers dropped to 4 states from 5 states after adding time inhomogeneity transition moving closer to a biologically and statistically interpretable level. Although adding the time inhomogeneous only reduce the number of states by 1, it does support that adding temporal heterogeneity reduce the number of states and at the same time, increase goodness of fit by BIC.

Among the different types of time inhomogeneous transition HMMs, sinusoidal time dependent transition HMMs yields the lowest BIC. The multiple block transitions was a good fit for the male panthers, using a combination of three transition matrix ($b_{1}$ for $t={1-6,21-24}$, $b_{2}$ for $ t={7-16}$, and $b_{3}$ for $t={17-20}$) and it reduced the number of states by 1 but had no effect on the females. The quadratic and sinusoidal time covariate transitions was motivated by the multiple block transition pattern attempting to capture the cyclic diurnal cycle and same complex level in terms of number of parameters. We concluded that sinusoidal time dependent transition HMM was suitable for modeling these panthers.

@@

Next, we used the fitted models and simulate the same number of observations as the observed movement data and compare the average distance by time with the observed data. Models that assume temporal homogeneity (FMM and HMM) fail to capture the diurnal activity cycle. Models incorporating temporal heterogeneity can capture the observed patterns in the hourly average step.

@@

Lastly, the observed data has strong 24 hour lag autocorrelation (correlation with itself) which suggest temporal heterogeneity. FMM has no autocorrelation at all. HMM without temporal heterogeneity can capture the first few lags of the ACF but is unable to capture lags beyond 7. FMM with temporal heterogeneity can capture lags beyond 12, but not short-term correlation. HMM with temporal heterogeneity can fully capture the autocorrelation pattern of the observed data.

<<state_plots, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="States">>=
temp5 <- tail(getpars(fitsin5s),10)
temp6 <- tail(getpars(fithomo6s),12)
dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
                     log_sd=temp5[c(2,4,6,8,10)], states="5") 
dat2 <- data.frame(log_mean=temp6[c(1,3,5,7,9,11)],
                     log_sd=temp6[c(2,4,6,8,10,12)],states="6") 
dat1 <- dat1[order(dat1$log_mean),]
dat2 <- dat2[order(dat2$log_mean),]

dat <- rbind(dat1,dat2)
ggplot(dat,aes(log_mean,log_sd,color=states))+
    geom_point()+geom_line()
@

@@

@@

\begin{table}[]
\centering
\caption{Number of Parameters for each model type. \textbf{BMB: supplementary material?}
\label{Cat-ID}
\begin{tabular}{cccc}
 Type & Model  & Number of States & Number of Parameters  \\
  FMM & FMM & 3 & 9    \\
  FMM & FMM & 4 & 12    \\
  FMM & FMM & 5 & 15   \\
  FMM & FMM & 6 & 18  \\
  FMM + TH & FMM + TH (Sin)  & 3 & 15    \\
  FMM + TH & FMM + TH (Sin)  & 4 & 20    \\
  FMM + TH & FMM + TH (Sin)  & 5 & 25    \\
  FMM + TH & FMM + TH (Sin)  & 6 & 30    \\
  HMM & HMM & 3 & 18    \\
  HMM & HMM & 4 & 28    \\
  HMM & HMM & 5 & 40   \\
  HMM & HMM & 6 & 54  \\
  HMM + TH & HMM + TH (Block)  & 3 & 36    \\
  HMM + TH & HMM + TH (Block)  & 4 & 60    \\
  HMM + TH & HMM + TH (Block)  & 5 & 90    \\
  HMM + TH & HMM + TH (Block)  & 6 & 126    \\
  HMM + TH & HMM + TH (Quad)  & 3 & 36    \\
  HMM + TH & HMM + TH (Quad)  & 4 & 60    \\
  HMM + TH & HMM + TH (Quad)  & 5 & 90    \\
  HMM + TH & HMM + TH (Quad)  & 6 & 126    \\
  HMM + TH & HMM + TH (Sin)  & 3 & 36    \\
  HMM + TH & HMM + TH (Sin)  & 4 & 60    \\
  HMM + TH & HMM + TH (Sin)  & 5 & 90    \\
  HMM + TH & HMM + TH (Sin)  & 6 & 126    \\
  HMM + TH & HMM + TH (Hourly)  & 3 & 225    \\
  HMM + TH & HMM + TH (Hourly)  & 4 & 396    
\end{tabular}
\end{table}

\bibliography{paper}

>>>>

\doublespacing
<<pkgs,message=FALSE,warning=FALSE,echo=FALSE>>=
library("ggplot2"); theme_set(theme_bw())
scale_colour_discrete <- function(...,palette="Set1")
    scale_colour_brewer(...,palette=palette)
scale_fill_discrete <- function(...,palette="Set1")
    scale_fill_brewer(...,palette=palette)
zmargin <- theme(panel.margin=grid::unit(0,"lines"))
library("knitr")
opts_chunk$set(fig.width=6,fig.height=4)
library("depmixS4")
library("plyr")
library("dplyr")
library("reshape2")
library("gridExtra")
@

@@

Figure 1 shows the BIC-optimal number of states for time homogenous models is consistent with \citeauthor{kerk2015hidden}'s results. Furthermore, BIC-optimal number of states does not depend on the model for both time-homogenous and time-inhomogenous within their respective framework, but time-inhomogenous models have the BIC-optimal number of states overall.

@@

As a complement, we also added FMM, FMM with sinusoidal temporal heterogeneities, and HMM with hourly transitions to compare the temporal effects in goodness of fit. Looking at Figure 2, models with temporal heterogenity is better than models without in both FMM and HMM frameworks, but time homogenous HMM models is better than FMM with sinusoidal temporal heterogeneity. Temporal heterogenity is good with good models such as the quadratic and sinuosidal. Multiple block models is good for cats with distinguishable diurnal patterns and hourly model is too complicated.

@@

Figure 3 shows a clear diurnal pattern from the average hourly step lengths from the observed data. Models that assume temporal homogeneity (FMM and HMM) fail to capture the diurnal activity cycle. Models incorporating temporal heterogeneity can capture the observed patterns without the Viterbi algorithm.

@@

The observed data has strong 24 hour lag autocorrelation (correlation with itself) which suggest temporal heterogeneity. HMM without temporal heterogeneity can capture short-term autocorrelation (without Viterbi algorithm) but is unable to capture lags beyond 7. FMM with temporal heterogeneity can capture lags beyond 12, but not short-term autocorrelations. HMM with temporal heterogeneity can fully capture the autocorrelation pattern of the observed data whereas the FMM completely missed the autocorrelations.

<<state_plots, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="States">>=
temp5 <- tail(getpars(fitsin5s),10)
temp6 <- tail(getpars(fithomo6s),12)
dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
                     log_sd=temp5[c(2,4,6,8,10)], Model="HMM + TH(sin)") 
dat2 <- data.frame(log_mean=temp6[c(1,3,5,7,9,11)],
                     log_sd=temp6[c(2,4,6,8,10,12)],Model="HMM") 
dat1 <- dat1[order(dat1$log_mean),]
dat2 <- dat2[order(dat2$log_mean),]

dat <- rbind(dat1,dat2)
sd1 <- ggplot(dat,aes(log_mean,log_sd,color=Model))+
    geom_point()+geom_line() + ggtitle("Cat 1")
@

<<state_plots cat2, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="States">>=
load("cat2sumdat.RData")
temp5 <- tail(getpars(fithomo5s),10)
temp4 <- tail(getpars(fitsin4s),8)
dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
                     log_sd=temp5[c(2,4,6,8,10)], Model="HMM") 
dat2 <- data.frame(log_mean=temp4[c(1,3,5,7)],
                     log_sd=temp4[c(2,4,6,8)],Model="HMM + TH(sin)") 
dat1 <- dat1[order(dat1$log_mean),]
dat2 <- dat2[order(dat2$log_mean),]

dat <- rbind(dat2,dat1)
sd2 <- ggplot(dat,aes(log_mean,log_sd,color=Model))+
    geom_point()+geom_line() + ggtitle("Cat 2")
@

<<state_plots cat14, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="States">>=
load("cat14sumdat.RData")
temp5 <- tail(getpars(fithomo5s),10)
temp4 <- tail(getpars(fitsin4s),8)
dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
                     log_sd=temp5[c(2,4,6,8,10)], Model="HMM") 
dat2 <- data.frame(log_mean=temp4[c(1,3,5,7)],
                     log_sd=temp4[c(2,4,6,8)],Model="HMM + TH(sin)") 
dat1 <- dat1[order(dat1$log_mean),]
dat2 <- dat2[order(dat2$log_mean),]

dat <- rbind(dat2,dat1)
sd14 <- ggplot(dat,aes(log_mean,log_sd,color=Model))+
    geom_point()+geom_line() + ggtitle("Cat 14")
@

<<state_plots cat15, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="States">>=
load("cat15sumdat.RData")
temp5 <- tail(getpars(fitsin5s),10)
temp6 <- tail(getpars(fithomo6s),12)
dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
                     log_sd=temp5[c(2,4,6,8,10)], Model="HMM") 
dat2 <- data.frame(log_mean=temp6[c(1,3,5,7,9,11)],
                     log_sd=temp6[c(2,4,6,8,10,12)],Model="HMM + TH(sin)") 
dat1 <- dat1[order(dat1$log_mean),]
dat2 <- dat2[order(dat2$log_mean),]

dat <- rbind(dat1,dat2)
sd15 <- ggplot(dat,aes(log_mean,log_sd,color=Model))+
    geom_point()+geom_line() + ggtitle("Cat 15")

grid.arrange(sd1,sd2,sd14,sd15)
@

@@

\begin{table}[]
\centering
\caption{Number of Parameters for each model type. 
\label{Cat-ID}
\begin{tabular}{cccc}
 Type & Model  & Number of States & Number of Parameters  \\
  FMM & FMM & 3 & 9    \\
  FMM & FMM & 4 & 12    \\
  FMM & FMM & 5 & 15   \\
  FMM & FMM & 6 & 18  \\
  FMM + TH & FMM + TH (Sin)  & 3 & 15    \\
  FMM + TH & FMM + TH (Sin)  & 4 & 20    \\
  FMM + TH & FMM + TH (Sin)  & 5 & 25    \\
  FMM + TH & FMM + TH (Sin)  & 6 & 30    \\
  HMM & HMM & 3 & 18    \\
  HMM & HMM & 4 & 28    \\
  HMM & HMM & 5 & 40   \\
  HMM & HMM & 6 & 54  \\
  HMM + TH & HMM + TH (Block)  & 3 & 36    \\
  HMM + TH & HMM + TH (Block)  & 4 & 60    \\
  HMM + TH & HMM + TH (Block)  & 5 & 90    \\
  HMM + TH & HMM + TH (Block)  & 6 & 126    \\
  HMM + TH & HMM + TH (Quad)  & 3 & 36    \\
  HMM + TH & HMM + TH (Quad)  & 4 & 60    \\
  HMM + TH & HMM + TH (Quad)  & 5 & 90    \\
  HMM + TH & HMM + TH (Quad)  & 6 & 126    \\
  HMM + TH & HMM + TH (Sin)  & 3 & 36    \\
  HMM + TH & HMM + TH (Sin)  & 4 & 60    \\
  HMM + TH & HMM + TH (Sin)  & 5 & 90    \\
  HMM + TH & HMM + TH (Sin)  & 6 & 126    \\
  HMM + TH & HMM + TH (Hourly)  & 3 & 225    \\
  HMM + TH & HMM + TH (Hourly)  & 4 & 396    
\end{tabular}
\end{table}

@@

<<adj_BIC_comparisons cat2, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall Adjusted BIC Comparisons between different HMM models">>=
load("cat2sumdat.RData")
ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ type,ncol=4)+
    scale_size_continuous( name="Num of Parameters") +
    geom_point(aes(size=parameters))+
    scale_colour_brewer(palette="Dark2")+
    labs(y=expression(paste(Delta, "BIC")),
         x="Number of States") +
    geom_line() + zmargin
@

<<adj_BIC_comparisons cat14, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall Adjusted BIC Comparisons between different HMM models">>=
load("cat14sumdat.RData")
ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ type,ncol=4)+
    scale_size_continuous( name="Num of Parameters") +
    geom_point(aes(size=parameters))+
    scale_colour_brewer(palette="Dark2")+
    labs(y=expression(paste(Delta, "BIC")),
         x="Number of States") +
    geom_line() + zmargin
@

<<adj_BIC_comparisons cat15, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall Adjusted BIC Comparisons between different HMM models">>=
load("cat15sumdat.RData")
ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ type,ncol=4)+
    scale_size_continuous( name="Num of Parameters") +
    geom_point(aes(size=parameters))+
    scale_colour_brewer(palette="Dark2")+
    labs(y=expression(paste(Delta, "BIC")),
         x="Number of States") +
    geom_line() + zmargin
@

<<avg_step_length_by_time cat2, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="Average step length by time of day">>=
load("cat2sumdat.RData")
  simdf <- simdat %>% dplyr:::select(-obs) %>% group_by(time) %>% summarise_each(funs(mean))
  obsdf <- simdat %>% dplyr:::select(c(obs,time)) %>% filter(!is.na(obs)) %>% group_by(time) %>% summarise_each(funs(mean)) %>% dplyr:::select(obs)
  avgdf <- cbind(simdf,obsdf)
  avgdf2 <- melt(avgdf,'time')
ggplot(avgdf2,aes(x=time,y=value,colour=variable)) + 
    geom_point() +geom_line()+ xlab("Time of Day") + ylab("Step lengths (log10)")+
    scale_x_continuous(breaks=0:23)
@

<<avg_step_length_by_time cat14, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="Average step length by time of day">>=
load("cat14sumdat.RData")
  simdf <- simdat %>% dplyr:::select(-obs) %>% group_by(time) %>% summarise_each(funs(mean))
  obsdf <- simdat %>% dplyr:::select(c(obs,time)) %>% filter(!is.na(obs)) %>% group_by(time) %>% summarise_each(funs(mean)) %>% dplyr:::select(obs)
  avgdf <- cbind(simdf,obsdf)
  avgdf2 <- melt(avgdf,'time')
ggplot(avgdf2,aes(x=time,y=value,colour=variable)) + 
    geom_point() +geom_line()+ xlab("Time of Day") + ylab("Step lengths (log10)")+
    scale_x_continuous(breaks=0:23)
@

<<avg_step_length_by_time cat15, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="Average step length by time of day">>=
load("cat15sumdat.RData")
  simdf <- simdat %>% dplyr:::select(-obs) %>% group_by(time) %>% summarise_each(funs(mean))
  obsdf <- simdat %>% dplyr:::select(c(obs,time)) %>% filter(!is.na(obs)) %>% group_by(time) %>% summarise_each(funs(mean)) %>% dplyr:::select(obs)
  avgdf <- cbind(simdf,obsdf)
  avgdf2 <- melt(avgdf,'time')
ggplot(avgdf2,aes(x=time,y=value,colour=variable)) + 
    geom_point() +geom_line()+ xlab("Time of Day") + ylab("Step lengths (log10)")+
    scale_x_continuous(breaks=0:23)
@

<<acf_plot cat2, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="ACF">>=
load('cat2sumdat.RData')
acffun <- function(sim){
  a <- acf(sim,na.action=na.pass,plot=FALSE)
  df <- data.frame(ACF=c(a$acf),lag=a$lag)
  return(df)
}

<<acf_plot cat14, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="ACF">>=
load("cat14sumdat.RData")
acffun <- function(sim){
  a <- acf(sim,na.action=na.pass,plot=FALSE)
  df <- data.frame(ACF=c(a$acf),lag=a$lag)
  return(df)
}

<<acf_plot cat15, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="ACF">>=
load("cat15sumdat.RData")
acffun <- function(sim){
  a <- acf(sim,na.action=na.pass,plot=FALSE)
  df <- data.frame(ACF=c(a$acf),lag=a$lag)
  return(df)
}

  temp <- simdat %>% dplyr:::select(-time)%>%ldply(.,acffun)
  ggplot(temp,aes(lag,ACF,colour=.id))+
    geom_point()+geom_line()
@
\bibliography{paper}
